The binary liquid model for gravity-driven microchannel simulations

### 2010-10-13 Text/paper.tex
More work on the paper.

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
This section is to introduce the procedure of nondimensionalization and
initialization techniques to be able to predict parameters necessary for
simulation. The capillary number governs the interface thickness and defined as:
\begin{equation}
Ca=\frac{U_{bubble} \mu_{liq}}{\gamma},
\end{equation}
where $\mu_{liq}$ is the liquid viscosity, $\gamma$ is the surface tension. The
bubble velocity $U_{bubble}$ is taken in the center of the bubble where the film
thickness is examined. 
While parameters as $\mu_{liq}$, $\gamma$ and capillary number $Ca$ are given
or defined through the model parameters, the velocity $U_{bubble}$ is not
defined. In the simulation we approximately match the velocity with the
Poiseuille profile center velocit, which is given by the equation:
\begin{equation}
\begin{aligned}
U_{bubble} \approx U_{liq}=\frac{H^2}{8
\mu_{liq}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{bubble}\approx
\frac{{N_y}^2}{\frac{8}{6}(\tau_{liq}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
By taking the Poiseuille profile one can estimate the body force through the
capillary number and grid resolution:
\begin{equation}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{N_y}^2}\gamma Ca.
\end{aligned}
\end{equation}
In reality we iterate the body force value
until the desired velocity and the capillary number as well are achieved.

@@

Here are two different examples presented that can quantitatively show the
initialization procedure for quite different capillary numbers. In these
examples we use $k=0.04$ and $A=0.04$ as optimal binary liquid model parameters.
That explicitly implies the surface tension to be $\sqrt{\frac{8 k
A}{9}}=0.0377$ and the characteristic length of the interface
$4\sqrt{\frac{k}{A}}=4$ lattice boltzmann units. For the sake of simplicity and
stability the relaxations parameter for liquid and gas phases are taken as
$\tau_{liq}=2.5$ and $\tau_{gas}=0.7$. That gives the gas-liquid viscosity ratio
as $10$.  
\begin{description}
 \item[I \bm{$Ca=0.005$}]
  The film width for such small Capillary numbers \cite{bretherton} is
given by equation:
  \begin{equation}
  \frac{h_{\infty}}{H}=1.3375 Ca^{2/3}=0.03899,
  \end{equation}
  where we assume that $2 H$ or $Ny$ in lattice units is the channel width.
Therefore one needs to resolve the interface which is around $2$ percent of
  whole channel width with for instance $8$ lattice units, where $4$ lattice
units occupies the interface width for given. Note that the interface occupies
$50$ percent of the interface.  Therefore the channel
  width is $400$ lattice units. We take the bubble length as $5$ channel widths
  and the distance between bubble is as $2$ bubble lengths. This will give
  right away the grid size as $400\mathrm{x}6000$. In terms of computer memory
requirement the calculations with the double precision accuracy demand:
\begin{equation}
\begin{aligned}
&2\text{ set of populations }\mathrm{x}\,\, 8\,\, \text{bytes per population
}\mathrm{x}\,\, 9 \text{ number of
populations}\\
&\mathrm{x}\,400\,\mathrm{x}\,6000\approx 345 \mathrm{MB}. 
\end{aligned}
\end{equation}
Even with this naive assumption for the $50$ percent resolution of the film
widht in comparison with the interface width, one can get the memory
requirement as $345 MB$.

@@

  The film width for such small Capillary numbers is taken from
  \ref{fig:heil:planar} for Reynolds number $40$ equals to $0.12$. We want to
  resolve $6\%$ of whole channel width with $12$ lattice units, where $4$
lattice units are related to the interface resolution.  Therefore the channel
  width is $200$ lattice units. We take the bubble length as $5$ channel widths
  and the distance between slugs we take as $2$ bubble lengths. This will give
  right away the grid size as $200\mathrm{x}3000$. The velocity can be
calculated from the Capillary number as:
  \begin{equation}
  U_{bubble}=0.05 \frac{\gamma}{\mu_{liq}}=0.05 \frac{0.037712}{0.6666}=2.82\,
  10^{-3}.
  \end{equation}
  The velocity implies the following pressure gradient calculated from the
  Poisuielle flow solution:
  \begin{equation}
  \begin{aligned}
  &\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{Ny^2}\gamma Ca=3.77\,10^{-7}.
  \end{aligned}
  \end{equation}
  The initialization approach is to take values from the already performed
numerical simulation and to calculate the grid numbers from them matching the
body force gradient through the solution of Poiseuille flow.

\end{description}
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/giavedoni_planar.eps}
\caption{\citet{giavedoni-numerical} gathered results across the
literature for different Capillary numbers \label{fig:giavedoni:planar}}
\includegraphics[width=0.97\textwidth]{Figures/heil-planar.eps}
\caption{\citet{heil-bretherton} performed simulations for different ranges of
Capillary numbers and Reynolds numbers. \label{fig:heil:planar}}
\end{figure}
Note that we base the initialization techniques through correlations for the
capillary number and the body force. However, in the binary liquid framework
those correlations give only the approximation. First of all the body force
which is based on the Poiesuille velocity profile for the liquid phase is the
biggest assumption. In reality the less viscous bubble will have a slippage
factor and its velocity is larger than the surrounding liquid velocity. By
plugging in the obtained velocity back to the capillary number equation the
larger capillary number can be obtained. While we present the initialization
tips in reality one needs to take the bubble velocity from the simulations and
recalculate all the necessary characteristics.

\subsubsection{The influence of initial film width. Easy parameters to check}

 Velocity profiles behind the bubble and in the front
  meniscus can be seen in Fig.
\ref{fig:velocity:profiles:different:widths:force}
  and \ref{fig:velocity:profiles:different:widths:pressure} for the force driven
  and pressure difference driven flows, correspondingly. As one can see average
  velocity for the bubble and the slug larger than the calculated velocity which
  should give larger the film. However, the film width for the planar case is
  different and can have another coefficient in front of the proportionality to
  $Ca^{2/3}$. Also, the phase gradient is taken as $0$ right now which can have
  influence on the measurements as well.
  \begin{figure}
  \includegraphics[width=0.47\textwidth]{Figures/vel_bubble_force_width_6.eps}
  \hfill
  \includegraphics[width=0.47\textwidth]{Figures/vel_bulk_force_width_6.eps}\\
  \includegraphics[width=0.47\textwidth]{Figures/vel_bubble_force_width_10.eps}
  \hfill
  \includegraphics[width=0.47\textwidth]{Figures/vel_bulk_force_width_10.eps}\\
  \includegraphics[width=0.47\textwidth]{Figures/vel_bubble_force_width_14.eps}
  \hfill
  \includegraphics[width=0.47\textwidth]{Figures/vel_bulk_force_width_14.eps}\\
  \caption{The velocity profiles in bulk and bubble for different initialization
  parameters as the
  bubble width $Ny-12$,$Ny-20$ and $Ny-28$.
  \label{fig:velocity:profiles:different:widths:force}}
  \end{figure}

\subsection{Grid refinement}
To examine the interface width we performed grid refinement for both force and
pressure difference driven flows. To avoid the effects of rear and front
menisci the problem is initialized with the film width occupying $12$ percent
of the channel height. All other parameters are kept the same, that includes
keeping the same Capillary number and viscosities ratio. Therefore the
following parameters were utilized: $Ny=51,100,149,198$,corresponding to the
effective channels widths as $H_{eff}=49,98,147,196$, $Nx=736,1471,2206,2941=15
H_{eff}$. It is easy to check
that keeping all liquid viscosities the same yields the following to be
constant while doing the grid refinement
$H^2\frac{\mathrm{d}P}{\mathrm{d}x}=Ny^2\frac{\mathrm{d}P}{\mathrm{d}x}=Const$.
The results for the force driven flows velocity phase profiles are shown in
Fig. \ref{fig:grid:profiles}. The corresponding ratios to the effective channel
width (without bounce-back nodes) is as $0.0809472165313,
0.0708111118485, 0.0678483852412, 0.0664499716904$. One can see
that for the channel widths as $149,198$, the results are convergent. The
velocities in the center of the bubble are $0.004176469553,
0.00401004133, 0.003969909065, 0.00394134433$, which corresponds
to the $Ca=\frac{\mu_{liq} U_{bubble}}{\gamma}=\frac{\frac{2}{3}\,
0.0045}{0.03771}=0.069$, which is slightly higher when for the assumed
profile. However, if we define Capillary number through the velocity of the
slug, then the Capillary number will have another value. The velocities of the
slug are $0.003411444831,0.003412693869,0.003420338754,0.003425757312$ and
Capillary number defined through these velocities is $0.0602$. Those numbers
are adequate if we assume that the results presented in
Fig.\ref{fig:giavedoni:planar} and Fig. \ref{fig:heil:planar} correspond to the
half of the channel but not the whole channel.

Let us calculate to which extent we need to refine the interface to obtain
convergent results. The interface itself occupies approximately $5 \xi$, where
$\xi=\sqrt{k/A}=1$. This interface width divided to the effective channel width
($5\xi/H_{eff}$) is as follows $0.102,0.051,0.034,0.025$ for the given
$H_{eff}$. However, the interface width needs to be takes as the ratio with the
film width which we want to resolve, this ratio $5\xi/H_{film}$ is as follows
$1.2605,  0.7205,  0.501,  0.3839$. In reality this number is
a bit overestimated as the interface doesn't fill the influence of the wall and
it maintains all the macroscopic parameters jumps. Therefore, one needs to
resolve the interface as $40\%$ of the expected film width.
\begin{figure}
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_49.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_98.eps}\\
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_147.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_196.eps}\\
\caption{Grid refined profiles for the effictive channel widths
$H_{eff}=49,98,147,196$.\label{fig:grid:profiles}}
\end{figure}

@@

\section{Capillary number region}
This section is devoted to the study of different Capillary number. We refer to
grid refinement section where we obtained that the interface width should be
at most $40$ percent of the interface width. For the benchmark results we
suggest the following Capillary numbers as:
\begin{equation}
0.03,0.05,0.08,0.1,0.2,0.4,0.6,0.8,1.0
\end{equation}
The corresponding approximate film widths from \cite{giavedoni-numerical} are as
follows (we take the half width as it's done in the grid refinement section):
$0.04,0.06,0.08,0.1,0.12,0.13,0.15$. $5$ lattice boltzmann units occupy
$40\%$ of the interface thickness. Therefore the grid size should be as
following $314\mathrm{x}4681$,$210\mathrm{x}3121$,
$158\mathrm{x}2341$, $127\mathrm{x}1876$, $106\mathrm{x}1561$,
$98\mathrm{x}1441$,$85\mathrm{x}1246$.

>>>>

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
This section is to introduce the procedure of nondimensionalization and
initialization techniques to be able to predict parameters necessary for
simulation. The capillary number governs the interface thickness and defined as:
\begin{equation}
Ca=\frac{U_{bubble} \mu_{liq}}{\gamma},
\end{equation}
where $\mu_{liq}$ is the liquid viscosity, $\gamma$ is the surface tension. The
bubble velocity $U_{bubble}$ is taken in the center of the bubble where the film
thickness is examined. 
While parameters as $\mu_{liq}$, $\gamma$ and capillary number $Ca$ are given
or defined through the model parameters, the velocity $U_{bubble}$ is not
defined. In the simulation we approximately match the velocity with the
Poiseuille profile center velocit, which is given by the equation:
\begin{equation}
\begin{aligned}
U_{bubble} \approx U_{liq}=\frac{H^2}{8
\mu_{liq}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{bubble}\approx
\frac{{N_y}^2}{\frac{8}{6}(\tau_{liq}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
By taking the Poiseuille profile one can estimate the body force through the
capillary number and grid resolution:
\begin{equation}
\label{poiseuille:velocity:center}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{N_y}^2}\gamma Ca.
\end{aligned}
\end{equation}
In reality we iterate the body force value
until the desired velocity and the capillary number as well are achieved.

@@

Here are two different examples presented that can quantitatively show the
initialization procedure for quite different capillary numbers. In these
examples we use $k=0.04$ and $A=0.04$ as optimal binary liquid model parameters.
That explicitly implies the surface tension to be $\sqrt{\frac{8 k
A}{9}}=0.0377$ and the characteristic length of the interface
$5\sqrt{\frac{k}{A}}=5$ lattice boltzmann units. For the sake of simplicity and
stability the relaxations parameter for liquid and gas phases are taken as
$\tau_{liq}=2.5$ and $\tau_{gas}=0.7$. That gives the gas-liquid viscosity ratio
as $10$.  
\begin{description}
 \item[I \bm{$Ca=0.005$}]
  The film width for such small Capillary numbers \cite{bretherton} is
given by equation:
  \begin{equation}
  \frac{h_{\infty}}{H}=1.3375 Ca^{2/3}=0.03899,
  \end{equation}
  where we assume that $2 H$ or $Ny$ in lattice units is the channel width.
Therefore one needs to resolve the interface which is around $2$ percent of
  whole channel width with for instance $8$ lattice units, where $5$ lattice
units occupies the interface width for given. Note that the interface occupies
$62.5$ percent of the interface.  Therefore the channel
  width is $400$ lattice units. We take the bubble length as $5$ channel widths
  and the distance between bubble is as $2$ bubble lengths. It gives the
grid size as $400\mathrm{x}6000$. In terms of computer memory
requirement the calculations with the double precision accuracy demand:
\begin{equation}
\begin{aligned}
&2\text{ set of populations }\mathrm{x}\,\, 8\,\, \text{bytes per population
}\mathrm{x}\,\, 9 \text{ number of
populations}\\
&\mathrm{x}\,400\,\mathrm{x}\,6000\approx 345 \mathrm{MB}. 
\end{aligned}
\end{equation}
Even with this naive assumption for the $50$ percent resolution of the film
widht in comparison with the interface width, one can get the memory
requirement as $345 MB$.

@@

  The film width for such small Capillary numbers is taken from
  \ref{fig:heil:planar} for Reynolds number $40$ equals to $0.12$. We want to
  resolve $6\%$ of whole channel width with $12$ lattice units, where $5$
lattice units are related to the interface resolution.  Therefore the channel
  width is $200$ lattice units. We take the bubble length as $5$ channel widths
  and the distance between slugs we take as $2$ bubble lengths. That gives
the grid size as $200\mathrm{x}3000$. The velocity can be
calculated from the Capillary number as:
  \begin{equation}
  U_{bubble}=0.05 \frac{\gamma}{\mu_{liq}}=0.05 \frac{0.037712}{0.6666}=2.82\,
  10^{-3}.
  \end{equation}
  The velocity implies the following pressure gradient calculated from the
  Poisuielle flow solution:
  \begin{equation}
  \begin{aligned}
  &\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{Ny^2}\gamma Ca=3.77\,10^{-7}.
  \end{aligned}
  \end{equation}
  The initialization approach is to take values from the already performed
numerical simulation and to calculate the grid numbers from them matching the
body force gradient through the solution of Poiseuille flow.

\end{description}
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/giavedoni_planar.eps}
\caption{\citet{giavedoni-numerical} gathered results across the
literature for different Capillary numbers \label{fig:giavedoni:planar}}
\includegraphics[width=0.97\textwidth]{Figures/heil-planar.eps}
\caption{\citet{heil-bretherton} performed simulations for different ranges of
Capillary numbers and Reynolds numbers. \label{fig:heil:planar}}
\end{figure}
Note that we base the initialization techniques through correlations for the
capillary number and the body force. However, in the binary liquid framework
those correlations give only the approximation. First of all the body force
which is based on the Poiesuille velocity profile for the liquid phase is the
biggest assumption. In reality the less viscous bubble will have a slippage
factor and its velocity is larger than the surrounding liquid velocity. By
plugging in the obtained velocity back to the capillary number equation the
larger capillary number can be obtained. While we present the initialization
tips, in reality one needs to take the bubble velocity from the simulations and
recalculate all the necessary characteristics or in other words adjust the
forcing to obtain the desired velocity/capillary number.

\subsection{Grid refinement}
To properly estimate the interface resolution one needs to perform the
convergence study dependance on the grid resolution. All parameters are fixed,
including a bubble velocity, the capillary number and change the grid
resolution. Our goal is to know exactly the ratio of the interface width to the
film thickness where results are not dependant on the grid resolution.

The example what we want to simulate is the capillary number
$0.05$. As it was pointed out the film thickness in that case is $6$ percent.
Note that in the case of half-way bounce-back walls which are used in the
simulations one needs to calculate the film thickness as: 
\begin{equation}
\delta=(\phi_0-0.5)/(N_y-2),
\end{equation}
where $\phi_0$ is the grid coordinate where phase field equals to $0$, $N_y-2$
is related to the effective channel height. Note that it's a big
simplifications to impose the boundary in the middle beetween the bounce-back
node and the fluid node. In the case of $\delta=0.06$ the initial grid number
is $N_y=102$, which gives the horizontal grid number as $N_x-1=15(N_y-2)=1500$.
The bubble is initialized as rectangular box with coordinates
$y=7\dots N_y-8$, $x=\frac{N_x}{3}\dots \frac{2 N_x}{3}$ and phase
$\phi_{bubble}=-1$. All other nodes are initialized with the phase field
$\phi=1$. The force gradient can be estimated through the Poiseuille
profile formula, Eq.\ref{poiseuille:velocity:center}, and it equals
$1.508\,10^{-6}$.

After choosing the reference parameters, the grid refinement needs to keep the
same the same macroscopic parameters as capillary number and viscosities ratio.
Therefore the simulations were performed with grid
numbers $Ny=102,127,152,177,202,227$, corresponding to the
effective channels widths as $H_{eff}=100,125,150,175$,
$Nx=1501,1876,2251,2625,3000,3375=15
H_{eff}$. It is easy to check
that keeping all liquid viscosities the same yields the following quantity to
be a
constant while doing the grid refinement
$H_{eff}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } =Const$.
{\color{red} Michal, please add some results here and description to it as well.
The results for the force driven flows velocity phase profiles are shown in
Fig. \ref{fig:grid:profiles}. The corresponding ratios to the effective channel
width (without bounce-back nodes) is as $0.0809472165313,
0.0708111118485, 0.0678483852412, 0.0664499716904$. One can see
that for the channel widths as $149,198$, the results are convergent. The
velocities in the center of the bubble are $0.004176469553,
0.00401004133, 0.003969909065, 0.00394134433$, which corresponds
to the $Ca=\frac{\mu_{liq} U_{bubble}}{\gamma}=\frac{\frac{2}{3}\,
0.0045}{0.03771}=0.069$, which is slightly higher when for the assumed
profile. However, if we define Capillary number through the velocity of the
slug, then the Capillary number will have another value. The velocities of the
slug are $0.003411444831,0.003412693869,0.003420338754,0.003425757312$ and
Capillary number defined through these velocities is $0.0602$. 
}
Let us calculate to which extent we need to refine the interface to obtain
convergent results. The interface itself occupies approximately $5 \xi$, where
$\xi=\sqrt{k/A}=1$. This interface width divided to the effective channel width
($5\xi/H_{eff}$) is as follows {\color{red} Please add new values here
$0.102,0.051,0.034,0.025$ } for the given
$H_{eff}$. Let us calculate the interface width ratio to the film thickness as
$5\xi/H_{film}$ is as follows
{\color{red} Please add here new values $1.2605,  0.7205,  0.501,  0.3839$}.
Therefore, one needs to
resolve the interface as {\color{red} Please give number here $40\%$} of the
expected film width for simulations to be converged.
{\color{red} If we need this pictures???
\begin{figure}
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_49.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_98.eps}\\
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_147.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/grid_phase_prof_196.eps}\\
\caption{Do we need these pictures? Grid refined profiles for the effictive
channel widths
$H_{eff}=49,98,147,196$.\label{fig:grid:profiles}}
\end{figure}
}

\section{Capillary number region}

@@

This section is devoted to the study of different Capillary number. We refer to
grid refinement section where we obtained that the interface width should be
at most $40$ percent of the interface width. For the benchmark results we
suggest the following Capillary numbers as:
\begin{equation}
0.03,0.05,0.08,0.1,0.2,0.4,0.6,0.8,1.0
\end{equation}
The corresponding approximate film widths from \cite{giavedoni-numerical} are as
follows (we take the half width as it's done in the grid refinement section):
$0.04,0.06,0.08,0.1,0.12,0.13,0.15$. $5$ lattice boltzmann units occupy
$40\%$ of the interface thickness. Therefore the grid size should be as
following $314\mathrm{x}4681$,$210\mathrm{x}3121$,
$158\mathrm{x}2341$, $127\mathrm{x}1876$, $106\mathrm{x}1561$,
$98\mathrm{x}1441$,$85\mathrm{x}1246$.
